### 2025-08-07 actuarial.qmd
20250807 - time as constraint

Another potential reason clinicians may not use actuarial judgment is time.
This can be a legitimate challenge to using actuarial models in some contexts.
Actuarial methods can take time to apply—they require assessing the releving criteria, putting the assessment scores into an actuarial formula, and interpreting the output score.
In some clinical situations, such as risk assessments in terms of whether someone is imminently suicidal or homicidal and how to proceed, clinicians must make immediate in-session decisions @Jacinto2018.
Thus, it can be challenging or impractical in some situations to conduct all of the relevant assessments using actuarial approaches.
Moreover, in some workplace settings, such as health care settings, clinicians may be pushed beyond their limits in terms of productivity expectations.
In these situations, one possible approach is to exercise their best possible judgment based on the available evidence [i.e., intuitive decision-making processes; @Jacinto2018], but to check it against an actuarial model after the fact.
It is valuable to evaluate to receive feedback on the accuracy of one's judgments and, when time permits, to evaluate one's intuitive decision-making against analytical decision-making.

>>>>

Another potential reason clinicians may not use actuarial judgment is time.
This can be a legitimate challenge to using actuarial models in some contexts.
Actuarial methods can take time to apply—they require assessing the relevant criteria, putting the assessment scores into an actuarial formula, and interpreting the output score.
In some clinical situations, such as risk assessments in terms of whether someone is imminently suicidal or homicidal and how to proceed, clinicians must make immediate in-session decisions [@Jacinto2018].
Thus, it can be challenging or impractical in some situations to conduct all of the relevant assessments needed to apply actuarial approaches.
Moreover, in some workplace settings, such as health care settings, clinicians may be pushed beyond their limits in terms of productivity expectations.
In these situations, one possible approach is to exercise one's best possible judgment in the moment based on the available evidence [i.e., intuitive decision-making processes; @Jacinto2018], but to check it against an actuarial model after the fact.
This is like comparing your decisions in a chess match, after the game ends, to how a computer would have moved in each of the situations.
It is valuable to receive feedback on the accuracy of one's judgments and, when time permits, to evaluate one's intuitive decision-making against analytical decision-making.
However, there are structured approaches to assessing suicide risk that are quick to administer [e.g., @Posner2011].

Nevertheless, the lack of time is perhaps less of a concern in the sports domain.
Teams have lots of data readily available on many players.
And, for player selection, teams have considerable time to implement statistical models of how players will perform to inform their selections.
There is less time during games to make in-game decisions and adjustments from complicated models; that said, teams are gaining greater adeptness with making in-game adjustments and decisions based on analytics.
Teams use actuarial charts that inform, for instance, [when to go for it on fourth down](#sec-coachingGoingForIt).
Now that many teams are motivated to use analytics to gain a leg up on their competitors, they are finding ways to make the time for it.
Clinicians and other medical experts, however, still have considerable room for improvement in terms of their adoption of actuarial approaches to judgment and prediction.
